Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-20 _tex/paper2.tex
Built site for gh-pages

Researchers would then proceed to identify the estimands.
\emph{Identification} is concerned with determining whether an estimator
can accurately compute an estimand based on its assumptions, regardless
of random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
\emph{estimator} is a method or function that transforms data into an
estimate \citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand. They are derived through \emph{estimation},
which refers to the process of integrating data with an estimator
\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow} provides a
visual representation of the process of transitioning from estimands to
estimates.

@@

\subsubsection{The general idea}\label{sec-appendix-B1}

@@

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. In contrast,
\emph{conditioning} involves restricting the focus to the subset of the
population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).\\

Note that if a student is assigned to intervention \(T_{i}=1\), the
potential outcome under \(T_{i}=2\) becomes a counterfactual, as it is
no longer observed nor observable. To address this challenge, the
structural approach extends the ICE to the \emph{average causal effect}
(ACE, Equation~\ref{eq-ACE}), representing the average difference
between observed potential outcomes and their counterfactual
counterparts.

\begin{equation}\phantomsection\label{eq-ACE}{
\tau = E[\tau_{i}] = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)]
}\end{equation}

Even though counterfactuals are unobserved and unobservable, researchers
can still identify the ACE from associational estimates by leveraging
the structural approach. The approach identifies the ACE by
statistically conditioning data on a \emph{sufficient adjustment set} of
variables \(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
This \emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. If such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}.

@@

\begin{equation}\phantomsection\label{eq-CACE}{
E[Y_{i} \:|\:T_{i}=t, X]
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mACE1}{
\begin{aligned}
  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

@@

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
models} that facilitate valid causal inference by guiding researchers in
determining which statistical models can identify an estimand (ACE,
CACE, or other), assuming the depicted causal structure is correct
\citep{McElreath_2020}.

SCMs and DAGs support causal inference through two key advantages.
First, regardless of complexity, they can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.
Figure~\ref{fig-IEflow} shows the role of theoretical models in the
inference process.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display these five fundamental building blocks. The left panels of the
figures show the formal mathematical models, represented by the SCMs,
defined in terms of a set of \emph{endogenous} variables
\(X=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. The right panels of the
figures display these DAGs. A DAG is a graph consisting of nodes
connected by edges, where the nodes represent random variables. The term
\emph{directed} means that the edges extend from one node to another,
with arrows indicating the direction of causal influence. The term
\emph{acyclic} implies that the causal influences do not form loops,
ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Although DAGs typically omit exogenous
variables for simplicity (the \emph{standard representation}), including
these variables, as demonstrated in the \emph{magnified representation}
in the figures, can be advantageous. Their inclusion helps to highlight
potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

@@

A careful examination of the figures highlights the theoretical
assumptions underlying the observed variables in these building blocks.
SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} depict two unconnected
nodes, representing a scenario where variables \(X_{1}\) and \(X_{3}\)
are independent or not causally related. SCM \ref{fig-scm_bb2} and DAG
\ref{fig-mdag_bb2} illustrate two connected nodes, representing a
scenario where a \emph{parent} node \(X_{1}\) exerts a causal influence
on a \emph{child} node \(X_{3}\). In this setup, \(X_{3}\) is considered
a \emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
are described as \emph{adjacent} because there is a \emph{direct path}
connecting. SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict a
\emph{chain}, where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\)
influences \(X_{3}\). In this configuration, \(X_{1}\) is a parent node
of \(X_{2}\), which is a parent node of \(X_{3}\). This structure
creates a \emph{directed path} between \(X_{1}\) and \(X_{3}\).
Consequently, \(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and
\(X_{2}\) fully \emph{mediates} the relationship between the two. SCM
\ref{fig-scm_bb4} and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork},
where variables \(X_{1}\) and \(X_{3}\) are both influenced by
\(X_{2}\). Here, \(X_{2}\) is a parent node that \emph{confounds} the
relationship between \(X_{1}\) and \(X_{3}\). Finally, SCM
\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} show a \emph{collider},
where variables \(X_{1}\) and \(X_{3}\) are concurrent causes of
\(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are not
causally related to each other but both influence \(X_{2}\) (an
\emph{inmorality}). Notably, all building blocks assume the errors are
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X1} \:\bot\:e_{X2}\),
\(e_{X1} \:\bot\:e_{X3}\), and \(e_{X2} \:\bot\:e_{X3}\).

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B1}. SCM \ref{fig-scm_example1}
and DAG \ref{fig-mdag_example1} depict the plausible causal structure
for this example. In this context, the variable \(X\) (socio-economic
status of the school) acts as a confounder in the relationship between
the teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\caption{\label{fig-example1}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B1}.}

@@

This encoding enables researchers with conceptual (theoretical)
knowledge in the form of an SCM and DAG to predict patterns of
(in)dependencies in the data. As highlighted by
\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
structure of these conceptual models without requiring the quantitative
details of the equations or the distributions of the errors. Moreover,
once researchers observe empirical data, the patterns of
(in)dependencies in this data can provide significant insights into the
validity of the proposed conceptual model.

The five fundamental building blocks described in
Section~\ref{sec-appendix-B21} clearly illustrate which (in)dependencies
can SMCs and DAGs predict. For instance, applying the BNF to the causal
structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
supports the theoretical assumption that the observed variables
\(X_{1}\) and \(X_{3}\) are unconditionally independent or d-separated
(\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\) \citep[pp.~24]{Neal_2020}.
Conversely, when \(X_{3}\) is unconditionally dependent on \(X_{1}\)
(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
Notably, these descriptions demonstrate the clear correspondence between
the structural equations illustrated in Section~\ref{sec-appendix-B21}
and the CPDs.

Beyond the insights gained from two-node structures, researchers can
uncover more nuanced patterns of (in)dependencies from chains, forks,
and colliders. These (in)dependencies apply to any data generated by
these structures. For instance, applying the BNF to the chain structure
depicted in the SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} allow
researchers to represent the joint distribution for the observed
variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). This expression
implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\).
Moreover, in the fork structure shown in the SCM \ref{fig-scm_bb4} and
DAG \ref{fig-mdag_bb4}, researchers can express the joint distribution
of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). This expression
allows researchers to further infer that \(X_{1}\) and \(X_{3}\) are
unconditionally dependent \((X_{1} \:\not\bot\:X_{3})\), but
conditionally independent when controlling for \(X_{2}\)
\((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally, researchers analyzing the
collider structure illustrated in the SCM \ref{fig-scm_bb5} and DAG
\ref{fig-mdag_bb5} can express the joint distribution of the observed
variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}, X_{3}) P(X_{3})\). This representation
allows researchers to infer that \(X_{1}\) and \(X_{3}\) are
unconditionally independent \((X_{1} \:\bot\:X_{3})\), but conditionally
dependent when controlling for \(X_{2}\)
\((X_{1} \:\not\bot\:X_{3} \:|\:X_{2})\). The authors \citet[pp.~37, 40,
41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
mathematical proofs for these conclusions.

Researchers can revisit the scenario in Section~\ref{sec-appendix-B1}
using these insights. By applying the BNF to the SCM
\ref{fig-scm_example2} and DAG \ref{fig-mdag_example2} structures, they
can represent the joint probability of the observed variables as
\(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this
expression, researchers can infer that the outcome \(Y\) is not
unconditionally independent of the teaching method \(T\)
\((Y \:\bot\:T)\). This dependency arises from two key structures: a
direct causal path from the teaching method \(T\) to the outcome \(Y\),
represented by the two-connected-nodes structure \(T \rightarrow Y\)
(black path), and a confounding non-causal path from the teaching method
\(T\) to the outcome \(Y\) through the socio-economic status of the
school \(X\), represented by the fork structure
\(T \leftarrow X \rightarrow Y\) (gray path).

These probabilistic insights also indicate that narrowing the focus to a
subset of the population defined by the socio-economic status of the
school \(X\), achieved by conditioning on \(X\) and depicted by a square
in DAG \ref{fig-mdag_example2}, effectively blocks all confounding
non-causal path introduced by the fork structure. Nevertheless, since
\(Y\) remains not d-separated from \(T\) even after conditioning
\((Y \:\not\bot\:T \:|\:X)\), this lack of d-separation can only be
attributed to the direct causal effect \(T \rightarrow Y\).

@@

\caption{\label{fig-example2}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B1}.}

@@

The structural approach to causal inference translate the probabilistic
insights from Section~\ref{sec-appendix-B23} into actionable strategies
for identifying the ACE from associational quantities. The approach
achieve this by relying on the modularity assumption and defining
interventions of the form \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
Section~\ref{sec-appendix-B1}). The \emph{modularity assumption} posits
that intervening on a node alters only the causal mechanism of that
node, leaving others unchanged \citep[pp.~34]{Neal_2020}.

The modularity assumption underpins the concepts of manipulated graphs
and Truncated Factorization, which are essential for representing
interventions within SCMs and DAGs. \emph{Manipulated graphs} simulate
physical interventions by removing specific edges from a DAG, while
preserving the remaining structure unchanged \citep[pp.~34]{Neal_2020}.
In parallel, \emph{Truncated Factorization} (TF) achieves a similar
simulation by removing specific functions from the conceptual model and
replacing them with constants, while keeping the rest of the structure
unchanged {[}Pearl\_2010{]}. The probabilistic implications of this
factorization are formalized in Equation~\ref{eq-trunc_factor}, where
\(S\) represents the subset of variables \(X_{p}\) directly influenced
by the intervention.

@@

Ultimately, backdoor adjustment enables researchers to express the ACE
as:

\begin{equation}\phantomsection\label{eq-backdoor_adjustment}{
\begin{aligned}
\tau &= E[\tau_{i}] \\
  & = E[X_{a} \:|\:do(X_{b}=1)]- E[X_{a} \:|\:do(X_{b}=2)] \\
  & = \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=1, X_{p}) P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=2, X_{p}) P(X_{p})
\end{aligned}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mACE2}{
\begin{aligned}
  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

>>>>

Researchers would then proceed to identify the estimands.
\emph{Identification} determines whether an estimator can accurately
compute an estimand based on its assumptions, regardless of random
variability \citep[pp.~4]{Schuessler_et_al_2023}. An \emph{estimator} is
a method or function that transforms data into an estimate
\citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand, and they are derived through
\emph{estimation}, which refers to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. The Identification-Estimation
flowchart \citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow}
provides a visual representation of the process of transitioning from
estimands to estimates.

@@

@@

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. Note that if a student
is assigned to intervention \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
observable. To address this challenge, the structural approach extends
the ICE to the \emph{average causal effect} (ACE,
Equation~\ref{eq-ACE}), representing the average difference between
observed potential outcomes and their counterfactual counterparts.

\begin{equation}\phantomsection\label{eq-ACE}{
\begin{aligned}
\tau & = E[\tau_{i}] \\
  & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)]
\end{aligned}
}\end{equation}

Even though counterfactuals are unobserved and unobservable, researchers
can still identify the ACE from associational estimates by leveraging
the structural approach. The approach identifies the ACE by
statistically conditioning data on a \emph{sufficient adjustment set} of
variables \(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
This \emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. If such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
\emph{conditioning} describes the process of restricting the focus to
the subset of the population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).

@@

\begin{equation}\phantomsection\label{eq-CACE}{
CACE_{t} = E[Y_{i} \:|\:T_{i}=t, X]
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mACE1}{
\begin{aligned}
  \tau & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)] \\
  & = E_{X}[CACE_{1} - CACE_{2}] \\
  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
models} that facilitate valid causal inference by guiding researchers in
determining which statistical models can identify an estimand (ACE,
CACE, or other), assuming the depicted causal structure is correct
\citep{McElreath_2020}. Figure~\ref{fig-IEflow} shows the role of
theoretical models in the inference process.

SCMs and DAGs support causal inference through two key advantages.
First, regardless of complexity, they can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.
Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
and Section~\ref{sec-appendix-B23} do so for the second.

@@

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display the five fundamental building blocks for SCMs and DAGs. The left
panels of the figures show the formal mathematical models, represented
by the SCMs, defined in terms of a set of \emph{endogenous} variables
\(X=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. The right panels of the
figures display these DAGs. A DAG is a graph consisting of nodes
connected by edges, where the nodes represent random variables. The term
\emph{directed} means that the edges extend from one node to another,
with arrows indicating the direction of causal influence. The term
\emph{acyclic} implies that the causal influences do not form loops,
ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Although the \emph{standard representation}
of DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} depicted in the figures offers some
advantages: including exogenous variables can help to highlight
potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

@@

A careful examination of the figures highlights the theoretical
assumptions underlying the observed variables in these building blocks.
SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} depict two unconnected
nodes, representing a scenario where variables \(X_{1}\) and \(X_{3}\)
are independent or not causally related. SCM \ref{fig-scm_bb2} and DAG
\ref{fig-mdag_bb2} illustrate two connected nodes, representing a
scenario where a \emph{parent} node \(X_{1}\) exerts a causal influence
on a \emph{child} node \(X_{3}\). In this setup, \(X_{3}\) is considered
a \emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
are described as \emph{adjacent} because there is a \emph{direct path}
connecting them. SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict
a \emph{chain}, where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\)
influences \(X_{3}\). In this configuration, \(X_{1}\) is a parent node
of \(X_{2}\), which is a parent node of \(X_{3}\). This structure
creates a \emph{directed path} between \(X_{1}\) and \(X_{3}\).
Consequently, \(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and
\(X_{2}\) fully \emph{mediates} the relationship between the two. SCM
\ref{fig-scm_bb4} and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork},
where variables \(X_{1}\) and \(X_{3}\) are both influenced by
\(X_{2}\). Here, \(X_{2}\) is a parent node that \emph{confounds} the
relationship between \(X_{1}\) and \(X_{3}\). Finally, SCM
\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} show a \emph{collider},
where variables \(X_{1}\) and \(X_{3}\) are concurrent causes of
\(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are not
causally related to each other but both influence \(X_{2}\) (an
\emph{inmorality}). Notably, all building blocks assume the errors are
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X1} \:\bot\:e_{X2}\),
\(e_{X1} \:\bot\:e_{X3}\), and \(e_{X2} \:\bot\:e_{X3}\).

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) acts as a confounder in the relationship between the
teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\caption{\label{fig-example1}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B}.}

@@

This encoding enables researchers with conceptual (theoretical)
knowledge in the form of an SCM and DAG to predict patterns of
(in)dependencies in the data. As highlighted by
\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
structure of these conceptual models without requiring the quantitative
details of the equations or the distributions of the errors. Moreover,
once researchers observe empirical data, the patterns of
(in)dependencies in the data can provide significant insights into the
validity of the proposed conceptual model.

The five fundamental building blocks described in
Section~\ref{sec-appendix-B21} clearly illustrate which variables'
(in)dependencies can SMCs and DAGs predict. For instance, applying the
BNF to the causal structure shown in the SCM \ref{fig-scm_bb1} and DAG
\ref{fig-mdag_bb1} supports the theoretical assumption that the observed
variables \(X_{1}\) and \(X_{3}\) are unconditionally independent
(\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\) \citep[pp.~24]{Neal_2020}.
Conversely, when \(X_{3}\) is unconditionally dependent on \(X_{1}\)
(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
Notably, these descriptions demonstrate the clear correspondence between
the structural equations illustrated in Section~\ref{sec-appendix-B21}
and the CPDs.

Beyond the insights gained from two-node structures, researchers can
uncover more nuanced patterns of variables' (in)dependencies from
chains, forks, and colliders. These (in)dependencies apply to any data
generated by these structures. For instance, applying the BNF to the
chain structure depicted in the SCM \ref{fig-scm_bb3} and DAG
\ref{fig-mdag_bb3} allow researchers to represent the joint distribution
for the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). This expression
implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\).
Moreover, in the fork structure shown in the SCM \ref{fig-scm_bb4} and
DAG \ref{fig-mdag_bb4}, researchers can express the joint distribution
of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). This expression
allows researchers to further infer that \(X_{1}\) and \(X_{3}\) are
unconditionally dependent \((X_{1} \:\not\bot\:X_{3})\), but
conditionally independent when controlling for \(X_{2}\)
\((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally, researchers analyzing the
collider structure illustrated in the SCM \ref{fig-scm_bb5} and DAG
\ref{fig-mdag_bb5} can express the joint distribution of the observed
variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}, X_{3}) P(X_{3})\). This representation
allows researchers to infer that \(X_{1}\) and \(X_{3}\) are
unconditionally independent \((X_{1} \:\bot\:X_{3})\), but conditionally
dependent when controlling for \(X_{2}\)
\((X_{1} \:\not\bot\:X_{3} \:|\:X_{2})\). The authors \citet[pp.~37, 40,
41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
mathematical proofs for these conclusions.

Then, researchers can revisit the scenario in
Section~\ref{sec-appendix-B} using these insights. In this context,
applying the BNF to the SCM \ref{fig-scm_example2} and DAG
\ref{fig-mdag_example2} structures, can help researchers to represent
the joint probability of the observed variables as \(P(Y, T, X) =\)
\(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this expression, researchers
can infer that the outcome \(Y\) is not unconditionally independent of
the teaching method \(T\) \((Y \:\bot\:T)\). This dependency arises from
two key structures: a direct causal path from the teaching method \(T\)
to the outcome \(Y\), represented by the two-connected-nodes structure
\(T \rightarrow Y\) (black path), and a confounding non-causal path from
the teaching method \(T\) to the outcome \(Y\) through the
socio-economic status of the school \(X\), represented by the fork
structure \(T \leftarrow X \rightarrow Y\) (gray path).

These probabilistic insights also indicate that narrowing the focus to a
subset of the population defined by the socio-economic status of the
school \(X\), achieved by conditioning on the variable and depicted by a
square in DAG \ref{fig-mdag_example2}, effectively blocks all
confounding non-causal path introduced by the fork structure.
Nevertheless, since \(Y\) remains dependent on \(T\) even after
conditioning \((Y \:\not\bot\:T \:|\:X)\), this dependency can only be
attributed to the direct causal effect \(T \rightarrow Y\).

@@

\caption{\label{fig-example2}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B}.}

@@

The structural approach to causal inference translate the probabilistic
insights from Section~\ref{sec-appendix-B23} into actionable strategies
for identifying the ACE from associational quantities. The approach
achieve this by relying on the modularity assumption and by re-defining
interventions in the form \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
Section~\ref{sec-appendix-B}). The \emph{modularity assumption} posits
that intervening on a node alters only the causal mechanism of that
node, leaving others unchanged \citep[pp.~34]{Neal_2020}.

The modularity assumption underpins the concepts of manipulated graphs
and Truncated Factorization, which are essential for representing
interventions within SCMs and DAGs. \emph{Manipulated graphs} simulate
physical interventions by removing specific edges from a DAG, while
preserving the remaining structure unchanged \citep[pp.~34]{Neal_2020}.
In parallel, \emph{Truncated Factorization} (TF) achieves a similar
simulation by removing specific functions from the conceptual model and
replacing them with constants, while keeping the rest of the structure
unchanged \citep{Pearl_2010}. The probabilistic implications of this
factorization are formalized in Equation~\ref{eq-trunc_factor}, where
\(S\) represents the subset of variables \(X_{p}\) directly influenced
by the intervention.

@@

Ultimately, the backdoor adjustment enables researchers to express the
ACE as:

\begin{equation}\phantomsection\label{eq-backdoor_adjustment}{
\begin{aligned}
\tau & = E[X_{a} \:|\:do(X_{b}=1)]- E[X_{a} \:|\:do(X_{b}=2)] \\
  & = E_{Xp}\left[ E[X_{a} \:|\:do(X_{b}=1), X_{p}]- E[X_{a} \:|\:do(X_{b}=2), X_{p}] \right] \\
  & = \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=1, X_{p}) \cdot P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=2, X_{p}) \cdot P(X_{p})
\end{aligned}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mACE2}{
\begin{aligned}
  \tau & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right] \\
  & = \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=2, X) \cdot P(X)
\end{aligned}
}\end{equation}
